UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-04-18 ms/main.tex
add some more maps. add citations, discussions

Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Analysis of the cluster by 23andMe suggested the Gujarati grouping is the result of endogamy~\cite{23andme}.

@@

As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}). The patterns in rural areas observed are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby.

\begin{figure}
    \centering
    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{A map of Great Britain coloured by a 3D UMAP projection. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (inset): Individuals who are closer to each other in the projection will appear to be closer in colour. Patterns in genetic similarity are visible in Scotland, South England, the East and West Midlands, and major urban centres. A flattened 2D view of of the 3D projection used for colouring is presented in the top right. To protect participant privacy, data has been randomized as explained in the materials and methods.}
    \label{fig:umap_ukbb_geo_map}
\end{figure}

@@

We tested different choices for perplexity in t-SNE. The default value of 30 provided comparable performance to other parameter choices.  Similarly, we tested different parameter choices for UMAP, with the clearest results generated by specifying 15 nearest neighbours (the default value) and a "minimum distance" between points in low dimensions of 0.5. UMAP developers described "sensible" values for nearest neighbours as between 5 and 50 and minimum distance between 0.5 and 0.001. Tuning these parameters will not change qualitative results much but may make patterns easier to identify. Increasing the number of neighbours will increase the computational load.

@@

To reduce the potential risks for re-identification from results in this publication, data has been randomly permuted so that the population characteristics are preserved but individual-level data is not presented directly in the figures. We rounded each attribute to an attribute-specific number of bins, and then permuted the data in the following way: 
For each point (i.e. each individual) in UMAP visualizations, and each attribute, we identified the 9 nearest neighbouring points, and copied the attribute from a randomly selected neighbor (thus allowing for the possibility of one value being printed twice).  Because this process is done independently for each visualization, a given point shown on the figure will copy values from different randomly selected individuals. Additionally, spatial coordinates have random noise added (normally distributed about 0 with a standard deviation of 50km) before binning to the nearest 50km.

For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}

@@

\subsubsection*{Acknowledgements}
We thank all participants in the HRS, UKBB, and 1KGP for providing their genetic data as well as the teams who generated and assembled the dataset. We also thank Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, and Audrey Grant for useful discussions about science, programming, and data access, and Selin Jessa for introducing us to UMAP. This research has been conducted using the UK Biobank Resource under Application Number 6728.

@@

@@

%%%%% boxplots end

>>>>

Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}, and analysis of the cluster by 23andMe suggests it is the result of endogamy\cite{23andme}.

@@

As an alternate visualization of geography and genetic diversity, we performed a 3D UMAP projection and converted the UMAP coordinates into RGB values, allowing us to plot individuals on a map of Great Britain, emphasizing both spatial gradients of genetic relatedness and increased diversity in urban centers (figure~\ref{fig:umap_ukbb_geo_map}). The patterns in rural areas observed are similar to those reported in~\cite{leslie2015fine} using the haplotype-based CHROMOPAINTER on British individuals whose grandparents lived nearby. Using data about country of birth, we also generalized this approach to colour a world map in figure~\ref{fig:umap_world_map}.

%\begin{figure}
%    \centering
%    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
%    \caption{A map of Great Britain coloured by a 3D UMAP projection. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (inset): Individuals who are closer to each other in the projection will appear to be closer in colour. Patterns in genetic similarity are visible in Scotland, South England, the East and West Midlands, and major urban centres. A flattened 2D view of of the 3D projection used for colouring is presented in the top right. To protect participant privacy, data has been randomized as explained in the materials and methods.}
%    \label{fig:umap_ukbb_geo_map}
%\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
    \label{fig:umap_ukbb_geo_map}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.9\columnwidth]{images/meanWorldMap.jpg}
   \caption{Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Vestiges of colonialism and related migrations are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
    \label{fig:umap_world_map}
      \end{subfigure}
      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods.}
\end{figure}

@@

We tested different choices for perplexity in t-SNE. The default value of 30 provided comparable performance to other parameter choices.  Similarly, we tested different parameter choices for UMAP, with the clearest results generated by specifying 15 nearest neighbours (the default value) and a "minimum distance" between points in low dimensions of 0.5. UMAP developers described "sensible" values for nearest neighbours as between 5 and 50 and minimum distance between 0.5 and 0.001. Tuning these parameters will not change qualitative results much but may make patterns easier to identify. Increasing the number of neighbours will increase the computational load, and a smaller minimum distance can break the connectivity between clusters, though the same individuals will continue to group together.

@@

Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate. For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country. Because many self-identified White British individuals were born abroad, including them everywhere would skew the colour scheme; they were included only if they were born in the UK, Europe, Australia, Canada, or the United States. This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations. For this reason, differences between Ireland/the UK and other countries appear exaggerated. Selecting a smaller minimum distance can result in breaking connections between clusters, leading to different colour groups appearing. We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}.

To reduce the potential risks for re-identification from results in this publication, data has been randomly permuted so that the population characteristics are preserved but individual-level data is not presented directly in the figures. We rounded each attribute to an attribute-specific number of bins, and then permuted the data in the following way: 
For each point (i.e. each individual) in UMAP visualizations, and each attribute, we identified the 9 nearest neighbouring points, and copied the attribute from a randomly selected neighbor (thus allowing for the possibility of one value being printed twice).  Because this process is done independently for each visualization, a given point shown on the figure will copy values from different randomly selected individuals. Additionally, spatial coordinates have random noise added (normally distributed about 0 with a standard deviation of 50km) before binning to the nearest 50km. For each point in figure~\ref{fig:umap_ukbb_geo_map} we identified the nearest 50 neighbouring individuals and copied the colour value from a randomly selected neighbour.}

@@

\subsubsection*{Acknowledgements}
We thank all participants in the HRS, UKBB, and 1KGP for providing their genetic data as well as the teams who generated and assembled the datasets. We also thank Jose Sergio Hleap, Mark Lathrop, Dominic Nelson, Markus Munter, Stephen Sawcer, and Audrey Grant for useful discussions about science, programming, and data access; David Poznik, Liz Babalola, and Adam Auton from 23andMe for discussing findings in the 1KGP; and Selin Jessa for introducing us to UMAP. This research has been conducted using the UK Biobank Resource under Application Number 6728.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanAsiaMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia.}
    \label{fig:supp_umap_ukbb_asia}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanCarribeanMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean.}
    \label{fig:supp_umap_ukbb_car}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanEuroMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe.}
    \label{fig:supp_umap_ukbb_eur}
\end{figure*}

@@

%%%%% boxplots end
\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanWorldMap_mindist001.jpg}
    \end{subfigure}
    \caption{Alternate colouring of figre~\ref{fig:umap_world_map} using a UMAP projection on the top 40 components with minimum distance 0.001.}
    \label{fig:supp_umap_ukbb_worldmap_alt}
\end{figure*}
